---
output:
  pdf_document:
    fig_caption: yes
header-includes: 
  - \renewcommand{\and}{\\}
  - \usepackage{float}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  out.width = "40%"
)
```

# Task 1

```{r calculations, include=FALSE}
read_adj_matrix <- function(path) {
  df <- read.csv2(path, row.names = 1, check.names = FALSE)
  mat <- as.matrix(df)
  storage.mode(mat) <- "numeric"
  mat[is.na(mat)] <- 0
  diag(mat) <- 0
  mat
}

suppressPackageStartupMessages(library(sna))
suppressPackageStartupMessages(library(network))

obsMat <- read_adj_matrix("Lintner/10_W1.csv")

n <- nrow(obsMat)
m <- sum(obsMat)
ndyads <- n * (n - 1) / 2

dc <- dyad.census(obsMat)
M <- as.numeric(dc[1, "Mut"])
A <- as.numeric(dc[1, "Asym"])
N <- as.numeric(dc[1, "Null"])

p <- m / (n * (n - 1))

pM <- M / ndyads
pA <- A / ndyads
pN <- N / ndyads

pM_ind <- p^2
pA_ind <- 2 * p * (1 - p)
pN_ind <- (1 - p)^2

d <- 9
p_out <- d / (n - 1)
p_mutual_out <- p_out^2

set.seed(161)
reps <- 5000

cugRec <- cug.test(obsMat, grecip, cmode = "edges", reps = reps)
cugInd <- cug.test(
  obsMat,
  centralization,
  cmode = "edges",
  FUN.arg = list(FUN = degree, cmode = "indegree"),
  reps = reps
)
cugTrans <- cug.test(obsMat, gtrans, cmode = "dyad.census", reps = reps)

recip_obs <- cugRec$obs.stat
indeg_cent_obs <- cugInd$obs.stat
trans_obs <- cugTrans$obs.stat

p_rec <- cugRec$pgteobs
p_ind <- cugInd$pgteobs
p_ind_lte <- cugInd$plteobs
p_trans <- cugTrans$pgteobs

ngte_rec <- sum(cugRec$rep.stat >= cugRec$obs.stat)
ngte_trans <- sum(cugTrans$rep.stat >= cugTrans$obs.stat)
```

## Data cleaning

Starting from the adjacency matrix in `Lintner/10_W1.csv`, the following
changes were made before analysis:

- All self-ties were removed (diagonal entries set to 0).
- Any missing entries were treated as absent ties (all `NA` values set to 0).

## Given

- Number of nodes: n = `r n`
- Directed ties (excluding diagonal): m = `r m`
- Dyad census: M = `r M` (mutual), A = `r A` (asymmetric), N = `r N` (null),
  total dyads = n(n-1)/2 = `r ndyads`

## (1) Tie probability

> Choosing an ordered pair \((i,j)\) uniformly at random with \(i \neq j\),
> there are \(n(n-1)\) possible directed ties.
>
> \[
> p = \frac{m}{n(n-1)} = \frac{`r m`}{`r n` \cdot `r (n - 1)`} =
> `r signif(p, 8)`
> \]

## (2) Dyad probabilities

> Choosing an unordered dyad \(\{i,j\}\) uniformly at random, there are
> \(n(n-1)/2\) dyads.
>
> \[
> p_M = \frac{M}{n(n-1)/2} = \frac{`r M`}{`r ndyads`} = `r signif(pM, 8)`
> \]
> \[
> p_A = \frac{A}{n(n-1)/2} = \frac{`r A`}{`r ndyads`} = `r signif(pA, 8)`
> \]
> \[
> p_N = \frac{N}{n(n-1)/2} = \frac{`r N`}{`r ndyads`} = `r signif(pN, 8)`
> \]

## (3) Dyad probabilities under tie independence with \(P(\text{tie}) = p\)

> If \(X_{ij}\) and \(X_{ji}\) are independent Bernoulli(\(p\)) within each
> dyad \(\{i,j\}\), then:
>
> \[
> p_M^{(ind)} = P(X_{ij}=1, X_{ji}=1) = p^2 = `r signif(pM_ind, 8)`
> \]
> \[
> p_A^{(ind)} = P(\text{exactly one tie}) = 2p(1-p) = `r signif(pA_ind, 8)`
> \]
> \[
> p_N^{(ind)} = P(\text{no ties}) = (1-p)^2 = `r signif(pN_ind, 8)`
> \]

## (4) Is tie independence reasonable here?

> No: compared to the independence model, the observed network has much more
> reciprocity (mutual dyads `r signif(pM, 4)` vs `r signif(pM_ind, 4)`) and
> many fewer asymmetric dyads (`r signif(pA, 4)` vs `r signif(pA_ind, 4)`).
>
> This suggests dyad-level dependence: if \(i\) nominates \(j\), then \(j\) is
> much more likely to nominate \(i\) than expected under independence.

## (5) Outdegree model with \(d=9\) (directional independence)

> If each node nominates exactly \(d=9\) friends and each potential receiver is
> equally likely, then for any ordered pair \(i \neq j\):
>
> \[
> P(i \to j) = \frac{d}{n-1} = \frac{`r d`}{`r (n - 1)`} = `r signif(p_out, 8)`
> \]
> \[
> P(\text{mutual}) = \left(\frac{d}{n-1}\right)^2 = `r signif(p_mutual_out, 8)`
> \]
>
> The mutual probability is close to the observed \(p_M\), but this model would
> imply \(m = n \cdot d = `r n * d`\) ties, which is larger than observed.

## (6) CUG tests

### (6.1) Hypotheses + conditioning

- `cug.test(obsMat, grecip, cmode="edges")` conditions on the number of edges
  (m fixed) and tests whether reciprocity is extreme.
- `cug.test(..., centralization, cmode="edges", FUN.arg=list(FUN=degree,
  cmode="indegree"))` conditions on the number of edges and tests whether
  in-degree centralization is extreme.
- `cug.test(obsMat, gtrans, cmode="dyad.census")` conditions on the dyad census
  (M, A, N fixed) and tests whether transitivity is extreme beyond what dyad-
  level reciprocity alone explains.

### (6.2) Results

#### Reciprocity (edges-conditioned)

> Observed reciprocity: `r signif(recip_obs, 6)`  
> One-sided p-value \(P(X \ge Obs)\): `r p_rec` (`r ngte_rec` / `r reps`)

```{r cug-recip-plot, echo=FALSE, fig.align='center', fig.cap="CUG test for reciprocity (cmode = edges)"}
plot(cugRec)
```

> Interpretation: With observed reciprocity `r signif(recip_obs, 4)` and
> \(P(X \ge Obs) = `r p_rec`\) (i.e., `r ngte_rec` out of `r reps` simulated
> graphs reached or exceeded the observed value), reciprocity is much higher
> than expected under an edges-only CUG null.

#### In-degree centralization (edges-conditioned)

> Observed in-degree centralization: `r signif(indeg_cent_obs, 6)`  
> Tail probabilities: \(P(X \ge Obs)\) = `r p_ind`, \(P(X \le Obs)\) = `r p_ind_lte`

```{r cug-indeg-plot, echo=FALSE, fig.align='center', fig.cap="CUG test for in-degree centralization (cmode = edges)"}
plot(cugInd)
```

> Interpretation: The observed in-degree centralization (`r signif(indeg_cent_obs, 4)`)
> is not extreme under an edges-only null (tail probabilities
> \(P(X \ge Obs) = `r p_ind`\), \(P(X \le Obs) = `r p_ind_lte`\)). This is
> consistent with “no strong evidence” of unusual centralization given \(m\).

#### Transitivity (dyad-census-conditioned)

> Observed transitivity: `r signif(trans_obs, 6)`  
> One-sided p-value \(P(X \ge Obs)\): `r p_trans` (`r ngte_trans` / `r reps`)

```{r cug-trans-plot, echo=FALSE, fig.align='center', fig.cap="CUG test for transitivity (cmode = dyad.census)"}
plot(cugTrans)
```

> Interpretation: With observed transitivity `r signif(trans_obs, 4)` and
> \(P(X \ge Obs) = `r p_trans`\) (i.e., `r ngte_trans` out of `r reps`
> simulated graphs reached or exceeded the observed value), transitivity is
> much higher than expected even after controlling for the dyad census (M, A, N).
