---
title: 'Task 1'
output: pdf_document
---

```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
library(ergm)
library(network)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# (1)
## i
The difference ($x_\text{after} - x_\text{before}$) is $\delta_\text{GWD}(x, \alpha) = (1 - e^{-\alpha})^k$.

## ii
```{r, echo=FALSE}
# Load required libraries
library(ggplot2)

# 1. Define the derived change statistic formula
calc_delta_gwd <- function(k, alpha) {
  return((1 - exp(-alpha))^k)
}

# 2. Set up parameters
k_max <- 10 # Adjust this to match the max out-degree of the Glasgow network
k_range <- 0:k_max 

# Vectors for decay values
alphas_neg <- c(-1, -0.5)
alphas_nonneg <- c(0, 0.1, 0.2, 0.3, 0.5, log(2), 1, 2, 3)

# 3. Create individual plots for NEGATIVE decay values

# Plot for alpha = -1
df_neg1 <- data.frame(k = k_range, delta = calc_delta_gwd(k_range, -1))
p_neg1 <- ggplot(df_neg1, aes(x = k, y = delta)) +
  geom_line(color = "red") + 
  geom_point(color = "red") +
  labs(title = expression(paste("GWD Change Statistic for ", alpha, " = -1")),
       x = "Current Out-Degree (k)",
       y = expression(delta[GWD])) +
  theme_minimal()
print(p_neg1)

# Plot for alpha = -0.5
df_neg05 <- data.frame(k = k_range, delta = calc_delta_gwd(k_range, -0.5))
p_neg05 <- ggplot(df_neg05, aes(x = k, y = delta)) +
  geom_line(color = "blue") + 
  geom_point(color = "blue") +
  labs(title = expression(paste("GWD Change Statistic for ", alpha, " = -0.5")),
       x = "Current Out-Degree (k)",
       y = expression(delta[GWD])) +
  theme_minimal()
print(p_neg05)

# 4. Create a single combined plot for NON-NEGATIVE decay values

# Create a data frame spanning all combinations of k and non-negative alphas
df_nonneg <- expand.grid(k = k_range, alpha = alphas_nonneg)
df_nonneg$delta <- calc_delta_gwd(df_nonneg$k, df_nonneg$alpha)

# Convert alpha to a factor so ggplot treats it as categorical data for coloring
# We round the labels slightly just to make log(2) look clean in the legend
df_nonneg$alpha_factor <- as.factor(round(df_nonneg$alpha, 3))

p_nonneg <- ggplot(df_nonneg, aes(x = k, y = delta, color = alpha_factor, group = alpha_factor)) +
  geom_line(linewidth = 0.8) + 
  geom_point(size = 1.5) +
  labs(title = "GWD Change Statistic for Non-Negative Decay Values",
       x = "Current Out-Degree (k)",
       y = expression(delta[GWD]),
       color = expression(alpha)) +
  theme_minimal() +
  theme(legend.position = "right")
print(p_nonneg)
```

## iii
- $\alpha = -1$ - The marginal effect is close to zero and then it explodes in absolute value (oscillating between positive and negative). For popular people, the effect is massive (in absolute value).
- $\alpha = -0.5$ - The marginal effect converges to zero oscillating between positive and negative. For popular people, the effect is very small.
- $\alpha \geq 0$ - The marginal effect is positive and goes to zero as $k$ grows. For popular people, the marginal effect is positive, but it decays quickly to zero. The greater the alpha, the slower the decay is.

## iv
We would like to use a value for $\alpha$ that is high (maybe $\alpha \geq 0.5$), so that the marginal effect for already active people is non-negligible.

# (2)
## i
```{r, eval=FALSE}
model1 <- ergm(glasgow_net1 ~ edges)
```

```{r, echo=FALSE}
model1 <- readRDS(file = "task1_model1.rds")
summary(model1)
```

## ii
### $\theta = -2$
```{r, eval=FALSE}
model2_neg2 <- ergm(glasgow_net1 ~ edges + offset(gwodegree(0.3, fixed = TRUE)), 
                      offset.coef = -2)
```
### $\theta = +2$
```{r, eval=FALSE}
model2_neg2 <- ergm(glasgow_net1 ~ edges + offset(gwodegree(0.3, fixed = TRUE)), 
                      offset.coef = +2)
```

## iii
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load required libraries
library(ergm)
library(ggplot2)
library(dplyr)
library(tidyr)

# Set a maximum out-degree to monitor (adjust if your observed network has higher degrees)
max_k <- 12 

# ==============================================================================
# 1. SIMULATE NETWORKS
# ==============================================================================
# We simulate 100 networks for each model.
# Note: While you can use simulate_formula(..., coef=coef(model)), using 
# simulate() directly on the fitted model objects inherits the formula and 
# coefficients automatically, which is safer and cleaner. 
# We use monitor = ~odegree(0:max_k) to track degree counts in the output.

# set.seed(123) # For reproducibility

model2_neg2 <- readRDS(file="task1_model2_neg2.rds")
model2_pos2 <- readRDS(file="task1_model2_pos2.rds")

# Model 1 (Edges only)
sim1 <- simulate(model1, nsim = 100, 
                   monitor = ~odegree(0:max_k), output = "stats")

# Model 2 (gwodegree coef = -2)
sim2_neg2 <- simulate(model2_neg2, nsim = 100, 
                       monitor = ~odegree(0:max_k), output = "stats")

# Model 2 (gwodegree coef = +2)
sim2_pos2 <- simulate(model2_pos2, nsim = 100, 
                       monitor = ~odegree(0:max_k), output = "stats")


# ==============================================================================
# 2. DATA WRANGLING
# ==============================================================================
# Helper function to extract the odegree columns and pivot the data for ggplot
format_stats <- function(stats_matrix, model_name) {
  df <- as.data.frame(stats_matrix)
  
  # Extract only the monitored out-degree columns
  odeg_cols <- grep("odegree", colnames(df), value = TRUE)
  df <- df[, odeg_cols]
  
  # Rename columns to strictly the numeric degree (e.g., "odegree0" -> "0")
  colnames(df) <- gsub("odegree", "", colnames(df))
  
  # Add simulation ID and reshape to long format
  df$Simulation <- 1:nrow(df)
  df_long <- pivot_longer(df, cols = -Simulation, 
                          names_to = "OutDegree", values_to = "Count")
  
  df_long$Model <- model_name
  df_long$OutDegree <- as.numeric(df_long$OutDegree)
  
  return(df_long)
}

# Format the data
df1 <- format_stats(sim1, "M1 (Edges Only)")
df2_neg2 <- format_stats(sim2_neg2, "M2 (gwodegree = -2)")
df2_pos2 <- format_stats(sim2_pos2, "M2 (gwodegree = +2)")


# ==============================================================================
# 3. SINGLE DISCRETE BAR-VIOLINS
# ==============================================================================
library(dplyr)
library(ggplot2)

# Helper function to create one unified "Bar-Violin" per comparison
plot_single_bar_violin <- function(df_ref, df_target, target_label, target_color) {
  
  # 1. Prepare the mirrored data
  # M1 counts become negative (left side), Target counts remain positive (right side)
  plot_data <- bind_rows(
      df_ref %>% mutate(Side = "M1 (Edges Only)"),
      df_target %>% mutate(Side = target_label)
    ) %>%
    group_by(Side, OutDegree) %>%
    summarise(MeanCount = mean(Count), .groups = 'drop') %>%
    mutate(DisplayCount = ifelse(grepl("M1", Side), -MeanCount, MeanCount))

  # 2. Plotting the single unified shape
  ggplot(plot_data, aes(x = OutDegree, y = DisplayCount, fill = Side)) +
    # width = 1 makes the discrete bars touch, forming a solid "violin" silhouette
    geom_col(width = 1, color = "white", linewidth = 0.1) + 
    coord_flip() + 
    # Use a single central axis at 0
    geom_vline(xintercept = 0, color = "black", linewidth = 0.5) +
    scale_fill_manual(values = setNames(c("gray70", target_color), 
                                        c("M1 (Edges Only)", target_label))) +
    # Make the X-axis labels (counts) absolute/positive
    scale_y_continuous(labels = abs) + 
    scale_x_continuous(breaks = 0:max_k) +
    labs(title = paste("Bar-Violin Comparison: M1 vs", target_label),
         subtitle = "Left: M1 (Edges) | Right: M2 (gwodegree)",
         x = "Out-Degree (k)", 
         y = "Mean Node Count",
         fill = "Model Side") +
    theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          # Center the plot around the spine
          axis.title.x = element_text(hjust = 0.5))
}

# --- GENERATE THE TWO COMPARISONS ---

# Comparison 1: M1 vs M2 (gwodegree = -2)
p_neg_violin <- plot_single_bar_violin(df1, df2_neg2, 
                                       "M2 (gwodegree = -2)", "#1f78b4")
print(p_neg_violin)

# Comparison 2: M1 vs M2 (gwodegree = +2)
p_pos_violin <- plot_single_bar_violin(df1, df2_pos2, 
                                       "M2 (gwodegree = +2)", "#e31a1c")
print(p_pos_violin)
```

## iv
The random network in M1 has an out-degree distribution that looks a lot like a normal distribution, which is expected as we only control for the number of edges.

### $\theta_\text{GWD} = -2$
Here the distribution is bimodal - out-degree values that are either $0$ (isolates) or "average" are more common than the other ones. The second mode is a bit higher than the mode in M1, since the expected density is fixed and these nodes "compensate" for the larger amount of isolates.

### $\theta_\text{GWD} = +2$
Here the distribution looks like a normal distribution, but much more narrow. That is because the network is rewarder for having higher degree values. The reason that the mode is not at a very high out-degree is that there are diminishing returns at $\alpha = 0.3$, so having a more "balanced allocation" is more preferable. 