---
title: 'Task 2'
output: pdf_document
---

```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
library(ergm)
library(network)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

f1_matrix <- as.matrix(read.csv("Glasgow/f1.csv", header = FALSE)) 
demographics <- read.csv("Glasgow/demographic.csv", header = TRUE)

# 3. Prepare the network object
# The description says "Pupils were asked to name...", indicating a directed network.
glasgow_net1 <- network(f1_matrix, directed = TRUE)

# Add the gender attribute to the network nodes
# Gender is coded as 1=boy, 2=girl
set.vertex.attribute(glasgow_net1, "gender", demographics$gender)

logdistance_matrix <- as.matrix(read.csv("Glasgow/logdistance.csv", header = FALSE))
alcohol <- read.csv("Glasgow/alcohol.csv", header = TRUE) 

```

# (1)

```{r, echo=FALSE}
model_1 <- readRDS(file = "task2_model_1.rds")
summary(model_1)
```

-   Negative (and significant) edge coefficient - networks tend to be sparse
-   Positive (and significant) homophily coefficient - networks tend to have homophily

```{r, echo=FALSE}
coefs <- coef(model_1)
theta_edges <- coefs["edges"]
theta_gender_match <- coefs["nodematch.gender"]

# Calculate the log-odds of a tie between two students of the SAME gender
# Change statistic for edges is 1, change statistic for matching gender is 1
log_odds_same_gender <- theta_edges + theta_gender_match

# Convert log-odds to probability
prob_same_gender <- exp(log_odds_same_gender) / (1 + exp(log_odds_same_gender))

# Print the probability
cat("Conditional probability of a tie between same-gender students:", prob_same_gender, "\n")
```

# (2)

```{r, eval=FALSE}
ergm_model_2_params <- glasgow_net1 ~ edges + 
  # Endogenous Hypotheses
  mutual + 
  gwesp(0.1, fixed = TRUE) + 
  
  # Exogenous Hypotheses: Gender
  nodeofactor("gender", levels = 1) + 
  nodeifactor("gender", levels = 1) + 
  
  # Exogenous Hypotheses: Alcohol
  nodeocov("alcohol") + 
  nodeicov("alcohol") + 
  absdiff("alcohol") +
  
  # Basic Exogenous Controls
  edgecov(logdistance_matrix) + 
  absdiff("age") +

  # Basic Endogenous Controls (Stabilizers)
  gwidegree(0.6, fixed = TRUE) + 
  gwodegree(0.6, fixed = TRUE)
noisy_model_2 <- ergm(
                ergm_model_2_params,
                # ERGMs with this many parameters can take a moment to compute
                control = control.ergm(
                  MCMC.samplesize = 5000, 
                  MCMC.interval = 8000, 
                  MCMC.burnin = 1 * 1000 * 1000, 
                  main.method = "Stochastic-Approximation",
                  
                  parallel = 8,           
                  parallel.type = "PSOCK"
                )
)
rough_estimates <- coef(noisy_model_2)
model_2 <- ergm(
  ergm_model_2_params,
  control = control.ergm(
    MCMC.samplesize = 5000, 
    MCMC.interval = 10 * 1000,
    MCMC.burnin = 100 * 1000, 
    init = rough_estimates,
    
    parallel = 8,           
    parallel.type = "PSOCK"
  )
)
```

```{r, echo=FALSE}
model_2 <- readRDS(file = "task2_model_2.rds")
```

# (3)

```{r, echo=FALSE}
library(coda)

# 2. Your model saves the data from all 8 cores in a list called "sample". 
# We use [[1]] to extract only the simulation data from the very first core.
single_thread_data <- model_2$sample[[1]]

# 3. Plot the trace and density for this single thread
plot(single_thread_data)
```

We see that the distributions are unimodal and mostly centered around 0. The traces oscillate around 0 without any patterns.

```{r, echo=FALSE}
# 2. Use the coda function 'autocorr()' to get the matrix of lag values
single_thread_lags <- autocorr(single_thread_data)

# 3. Print the results to your console
print(single_thread_lags)
```

From the lags we see that the model explores the state space very well, since autocorrelation becomes very close to 0.

# (4)

```{r, eval=FALSE}
model_2_gof <-  gof(
  model_2,
  GOF = ~ idegree + odegree + espartners + distance
)
```

```{r, echo=FALSE}
model_2_gof <- readRDS(file = "task2_model_2_gof.rds")
model_2_gof
```
From the summary table we see that the p-values are high (not all of them are significant on the 5%) level. This also indicates that the model fits well based on the statistics that it is trained on.

```{r, echo=FALSE}
ergm:::plot.gof(model_2_gof)
```
Plotting the auxiliary statistics shows us that the model explores networks with similar in- and out-degree distributions, but the explored networks are much less connected than the reference one. This can be seen in the `distance` and `espartners` diagrams.

# (5)

```{r, echo=FALSE}
summary(model_2)
```

-   i - reciprocity is supported by data
-   ii - transitivity (gwesp.OTP.fixed.0.1) is supported by the data
-   iii, iv - cannot tell as the coefficients are insignificant
-   v - the opposite is supported
-   vi - supported
-   vii - supported (negative coefficients, so the abs difference should be small)

# (6)
????

# (7)
- **H1** - children with similar literacy scores tend to become friends (literacy score homophily)
- **H2** - Girls tend to receive more friend nominations than they send
- **H3 (a bit unsure about this one)** - alcohol homophily is more prevalent in boys compared to girls
