---
title: 'Task 3'
output: pdf_document
---

```{r, echo=FALSE, message=FALSE}
library(RSiena)
source("printSiena.R")
```

# (1)
## (1.1)
- Jaccard 1->2: 0.3036496
- Jaccard 2-3: 0.3506686

## (1.2)
```{r, eval=FALSE}
# 1. Create the initial effects object based on your data
myeff <- getEffects(mydata)

# 2. Add Endogenous Controls (Network Structure)
# Note: outdegree (density) and recip (reciprocity) are already included by default
myeff <- includeEffects(myeff, transTrip, cycle3)

# 3. Add Exogenous Controls (Demographics)
# Controlling for gender dynamics
myeff <- includeEffects(myeff, egoX, altX, sameX, interaction1 = "gender")

# Controlling for base alcohol tendencies (sender and receiver effects)
myeff <- includeEffects(myeff, egoX, altX, interaction1 = "alcohol_covar")
# (Optional) You can also add age controls if desired:
# myeff <- includeEffects(myeff, simX, interaction1 = "age")

# 4. Add Hypothesis Tests!
# H(i): Popularity
myeff <- includeEffects(myeff, inPop)

# H(ii): Alcohol Similarity
myeff <- includeEffects(myeff, simX, interaction1 = "alcohol_covar")

# H(iii): Distance / Proximity
myeff <- includeEffects(myeff, X, interaction1 = "distance")
```

## (1.3)
```{r, echo=FALSE}
model0 <- readRDS("task3_model0.rds")
printSiena(model0)
```

```{r}
t.conv <- apply(model0$sf, 2, mean) / apply(model0$sf, 2, sd)
t.conv
overall <- sqrt(t(apply(model0$sf, 2, mean)) %*% solve(cov(model0$sf)) %*% apply(model0$sf, 2, mean))
overall
```

We can see that the model converges.

## (1.4)
```{r, echo=FALSE}
gofEvId <- readRDS("task3_model0_gofEvId.rds")
gofEvOd <- readRDS("task3_model0_gofEvOd.rds")
gofEvGD <- readRDS("task3_model0_gofEvGD.rds")
gofEvTC <- readRDS("task3_model0_gofEvTC.rds")
plot(gofEvId)
plot(gofEvOd)
plot(gofEvGD)
plot(gofEvTC, center = TRUE, scale = TRUE)
```

We can see that the model has good GOF in terms of in-degree distribution, but everything else is quite poor. This means that the model does not represent the data well.

**From now on (for interpretation purposes), we pretend that the model has good GOF, so we can draw conclusions.**

## (1.5)
- i - The coefficient is negative and significant. We reject the hypothesis, since we observe the opposite effect.
- ii - positive and significant - hypothesis is supported
- iii - the coefficient is negative and significant - hypothesis is supported

# (2)

## (2.1)
```{r, eval=FALSE}
# =====================================================================
# SELECTION PART (Network Evolution)
# =====================================================================
# Endogenous Controls
myeff <- set_effect(myeff, list(transTrip, cycle3), depvar="friendship")

# Exogenous Controls (Gender)
myeff <- set_effect(myeff, list(egoX, altX, sameX), covar1="gender", depvar="friendship")

# Hypothesis (i): Popularity -> Friendship
myeff <- set_effect(myeff, list(inPop), depvar="friendship")

# Hypothesis (ii): Alcohol Similarity -> Friendship (Homophily)
# Note: covar1 is now targeting our new behavior variable "alcohol_beh"
myeff <- set_effect(myeff, list(simX, egoX, altX), covar1="alcohol_beh", depvar="friendship")

# Hypothesis (iii): Distance -> Friendship
myeff <- set_effect(myeff, list(X), covar1="distance", depvar="friendship")


# =====================================================================
# INFLUENCE PART (Behavior Evolution)
# =====================================================================
# Note: RSiena automatically includes the 'linear' and 'quad' (shape) 
# effects for behavior, which act as your basic endogenous controls.

# Exogenous Control: Gender's effect on alcohol consumption
# (It's good practice to control for basic demographics in behavior too)
myeff <- set_effect(myeff, list(effFrom), covar1="gender", depvar="alcohol_beh")

# Hypothesis (iv): Popularity -> Alcohol
# Does indegree in the friendship network predict alcohol behavior?
myeff <- set_effect(myeff, list(indeg), covar1="friendship", depvar="alcohol_beh")

# Hypothesis (v): Assimilation / Social Influence
# Do students adjust alcohol consumption to be similar to their friends?
myeff <- set_effect(myeff, list(avSim), covar1="friendship", depvar="alcohol_beh")
```

```{r, echo=FALSE}
modelCoev <- readRDS("task3_modelCoev.rds")
```

## (2.2)
```{r}
t.conv <- apply(model0$sf, 2, mean) / apply(model0$sf, 2, sd)
t.conv
overall <- sqrt(t(apply(model0$sf, 2, mean)) %*% solve(cov(model0$sf)) %*% apply(model0$sf, 2, mean))
overall
```
The model converges.

## (2.3)
```{r}
gofCoevId <- readRDS("task3_modelCoev_gofCoevId.rds")
plot(gofCoevId)

# Outdegree distribution
gofCoevOd <- readRDS("task3_modelCoev_gofCoevOd.rds")
plot(gofCoevOd)

# Triad census
gofCoevTC <- readRDS("task3_modelCoev_gofCoevTC.rds")
plot(gofCoevTC)

# Geodesic distance
gofCoevGD <- readRDS("task3_modelCoev_gofCoevGD.rds")
plot(gofCoevGD)

# Behaviour distribution
gofCoevBeh <- readRDS("task3_modelCoev_gofCoevBeh.rds")
plot(gofCoevBeh)

```

GOF of in-degree distribution is ok, the rest is not good. The model does not represent the data very well.

**We (again) assume that the model does assume the data, so we can interpret what we have**

## (2.4)
```{r}
printSienaCoev(modelCoev)
```

- iv - coefficient is insignificant. cannot confirm hypothesis.
- v - coefficient is positive and significant - confirm hypothesis

## (2.5)
We see both selection and influence effects:
- Pupils with similar alcohol consumption become friends, but also friends change their alcohol consumption to be similar

We also notice that these selection processes are also in place:
- Students become friends with other popular students
- Reciprocity

## (3)
- We can add `outActSqrt`. It is closely related to the out-degree distribution and should help us steer the model in the right direction
- We can add `gwespFF` to control for the amount of shared partners. This will help us improve the distance distribution.

## (4)
- Students with many friends tend to increase or maintain alcohol consumption
- Students tend to adjust their literacy score results towards the ones of their friends
